TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-10-16 manuscript/manuscript.Rmd
Rewording results

Network properties, such as connectivity (aka degree distribution) [@Ghiassian:2015aa; @Lee:2008aa; @Barabasi:2004aa; @Feldman:2008aa], can be compared to better understand the relationships among phenotypes [@Vidal:2011aa; @Barabasi:2011aa; @Wang:2011aa]. Networks formed by phenotype-associated genes, or, more precisely, their protein-protein products, are thought to have high connectivity [@Huang:2006aa; @Krauthammer:2004aa; @Goh:2007aa]. This intuition reflects a well-known "guilt-by-association" principle that genes (or, more precisely, their products) that form an interaction network are more likely to share similar functions [@Farber:2013aa], or co-expression patterns [@Huang:2006aa; @Michalak:2008aa]. Although this view has been criticized [@Gillis:2012aa], several studies consistently observed phenotype-associated genes to be either connected as a single network or to participate in common phenotype-specific subnetworks [@Menche:2015aa; @Lee:2011aa; @Akula:2011aa; @Ideker:2008aa; @Iossifov:2008aa]. Therefore, high connectivity among phenotype-specific genes may indicate coherent molecular mechanisms that could be targeted therapeutically. In terms of graph theory, we are asking whether a phenotype-specific network is a community - a highly connected subgraph relatively well-separated from the rest of the network [@fortunato2010community]. However, the direct comparison of connectivity across phenotype-specific networks is hindered by the fact that different phenotypes are associated with different numbers of genes forming networks of different sizes and configurations.

<!--This study compares the connectivity of networks formed by disease- or trait-associated genes and gene products. We hypothesized that genes associated with (categories of) diseases share common molecular mechanisms and, hence, form connected (i.e., coherent) networks. Therefore, the level of connectivity may inform us about similarities and differences among the diseases. The terms “network connectivity,” “coherence,” or “average degree distribution” are used interchangeably as appropriate. We developed a method for normalizing the internal network connectivity that accounts for differences in network sizes formed by disease-associated genes. We compared the slope of normalized internal vs. external network connectivity between psychiatric, neurologic, autoimmune, cancer, cardiovascular, metabolic and other diseases and traits. We used KEGG pathways [30] as a reference for a coherent network with common molecular mechanisms, and we used sets of randomly sampled genes as a reference for the absence of coherence. We found that genes associated with metabolic diseases and traits and autoimmune diseases formed coherent networks similar to KEGG pathways. On the contrary, the coherence of networks of psychiatric and neurologic disease-associated genes, as well as intelligence-associated traits, was similar to that of randomly selected genes. Cancer, anthropometric traits, eye-, and cardiovascular diseases and traits formed networks of intermediate connectivity. Our study identifies significant differences between categories of diseases in terms of network properties of the disease-associated genes and highlights the limitations of the current understanding of the molecular determinants of psychiatric disorders.-->
This study presents a novel measure of network coherence as a slope of the internal vs. external degree distributions. The internal degree distribution is defined as gene-specific interaction counts within a protein-protein interaction (PPI) network formed by phenotype-associated genes, while the external degree distribution counts interactions with other genes in the whole PPI network. We used selected gene sets from the MSigDb database [@Subramanian:2005aa] as a reference for highly coherent networks, while sets of randomly sampled genes were used as the reference for the absence of coherence. Using MSigDb and random networks matched in size to phenotype-specific networks, we derived a normalized measure of coherence that can be compared across phenotypes. We hypothesized that the level of coherence may inform us about similarities and differences among phenotypes. Using two PPI databases (STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa]), we compared coherence of 133 phenotypes from GWAScatalog [@Hindorff:2009aa; @MacArthur:2017aa]. Our results show the tendency of complex phenotypes, such as intelligence-related traits, to have low coherence and metabolic traits and diseases to have high coherence. Our method enables direct comparison of coherence measures and highlights gaps in the current understanding of molecular mechanisms of many phenotypes, e.g., Major Depressive Disorder having the lowest coherence in the already low-coherence "Psychiatric disease" category.

@@

To compare coherences, we measured coherence as the slope of a line through a scatterplot of internal vs. external degree distributions. That is, we visualized internal (X-axis) vs. external (Y-axis) degrees for each phenotype-specific gene on a single plot and fit a regression line through it, enforcing the fit through the origin. An example of a network with the highest coherence (full internal connectivity, zero external connections) would be represented by a horizontal line. Conversely, an example of a network with the lowest coherence would be a set of non-interacting genes (zero internal connections and full external connectivity) would be represented by a vertical line. Consequently, slopes of internal vs. external degree distributions of phenotype-specific networks would represent the corresponding levels of coherence. In summary, comparing internal vs. external degree distributions represents a viable metric to quantify and compare the molecular coherence of phenotype-associated networks.

@@

Direct comparison of degree distributions of the networks formed by phenotype-specific genes is hindered by the fact they depend on network size and configuration [@Leskovec:2010:ECA:1772690.1772755]. An intuitive example is to consider a ring and a fully connected network of size 10 and 20 nodes, respectively (Figure 1). The degree distributions of ring networks can be directly compared regardless of network size (vectors of "2" of length 10 and 20, respectively, Figure 1A).  Although the fully connected networks are similar in that they are fully connected (i.e., similarly coherent), their degree distributions differ (all nodes in the 10-node network have a degree "9", while all nodes in the 20-node network have degree "19", Figure 1B). <!--Normalizing the degree distributions of the fully connected networks by dividing them by network size makes them identical as would be expected for networks with similar coherence. However, similar normalization of ring networks will make their degree distributions different (0.22 for the 10-node network and 0.11 for the 20-node network), leading to the illusory impression of differences in coherence between intuitively equally coherent networks. -->These observations highlight the difficulty in comparing degree distributions of phenotype-associated gene networks due to two facts: 1) network sizes differ, and 2) network configuration (fully connected, ring, or intermediate connectivity) is unknown.

@@

Given the dependence of degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we created size-matched references of high and low coherence. Specifically, we selected 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections) and 100 random networks matched in size to the corresponding network of phenotype-specific genes. Consequently, we use median slopes of internal vs. external degree distributions created by these size-matched references to normalize the slopes of phenotype-specific networks to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed alleviates the dependency between coherence and network size (Average PCC = 0.01, Supplementary Figure S1) and allows for direct comparison of normalized coherences across phenotypes.

@@

Besides size dependency, coherence estimates depend on the choice of the PPI database used to build networks and estimate the corresponding internal vs. external degree distributions. To evaluate the consistency of coherence estimates, we assessed the pairwise correlation of normalized coherences obtained using different PPI databases. Correlation between coherence estimates obtained using "STRING" and "STRING filtered" was high (PCC = 0.80, Supplementary Figure S1). Expectedly, coherence estimates using Biogrid database were less similar to that of "STRING" and "STRING filtered"; however, the overall correlation remained high. Notably, coherence estimates using "STRING filtered" database were most similar to that of "Biogrid" (PCC = 0.58), indicating that filtering step indeed removes "noisy" PPIs and improves quality of coherence estimates. Overall, these results demonstrate consistency in coherence estimates using different PPI databases.

@@

We estimated coherence of 49 traits from five categories (Figure 3, Supplementary Table S3, Supplementary Figures S2-S3). Among anthropometric traits, "Weight" and "Body Mass Index" had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, metabolic trait category had the highest overall coherence (mean coherence 0.86, Table 1), and these results were consistent when using other PPI databases (Supplementary Table S1). Several notable examples of low coherence include "Obesity-related traits" and "Height" (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the "Intelligence" category had low overall coherence (mean coherence = 0.41). These results highlight the relatively good knowledge of genetics of metabolic traits which corresponds to highly coherent networks. Conversely, current knowledge about the genetics of intelligence-related traits appears insufficient as reflected in the low coherence of the corresponding networks.

@@

We further estimated coherence of 67 diseases from seven categories (Figure 4, Supplementary Figures S4-S5, Supplementary Table S3). Metabolic diseases had high overall coherence (average coherence 0.91, Table 1), with "Obesity" and "Type 2 diabetes" being the most coherent (1.01 and 0.91, respectively). Similarly, the average coherence for cancer diseases was 0.72, with "Testicular germ cell tumor" and "Pancreatic cancer" being highly coherent (0.92 and 0.74, respectively). The high coherence of metabolic and cancer disease networks highlight the relatively good knowledge about their genetic component.

@@

"Rheumatoid arthritis", "Systemic Lupus Erythematosus", and "Systemic sclerosis" were the most coherent in the "Autoimmune" category (0.86, 0.80, 0.79, respectively). On the other end of the coherence spectrum, "Allergic rinitis" and "Atopic dermatitis" were the least coherent (0.51, 0.60, respectively). The average coherence of autoimmune diseases was high, 0.70 (Table 1), with all autoimmune diseases, except "Allergic rinitis", having coherence estimates > 0.60 (Supplementary Table S3). Diseases in the "Neurologic" category were similarly highly coherent (average coherence 0.72). However, the average coherence of neurologic disorders using STRING database was 0.49, suggesting that, with larger number of networks being estimated the overall coherence of neurologic diseases is low. "Alzheimer's disease" and "Parkinson's disease" were the most coherent (0.88 and 0.84, respectively), while "Migraine" was the least coherent (0.54). Diseases in the "Eye disease" category were more heterogeneous, with coherence ranging from 0.35 for "Corneal astigmatism" to 1.01 for "Refractive error". The average coherence of diseases in the "Eye disease" category was relatively high (0.66). These results suggest good overall knowledge of the genetics of autoimmune, neurologic, and eye-related diseases and highlight limited knowledge for some diseases in these categories.

+-------+
| ![](Figures/Fig4_summary_diseases.png)      |
| **Figure 4. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).   |
+-------+

@@

Diseases in "Psychiatric" and "Cardiovascular" categories showed the lowest average level of coherence (0.64 and 0.66, respectively, Table 1). The corresponding coherence estimates were highly heterogeneous. Among cardiovascular diseases, "Coronary artery disease" showed the highest level of coherence (0.83), while "Hypertension" had the lowest coherence (0.53). Among psychiatric disorders, "Smoking behavior" and "Alcohol dependence" had high coherence (0.87 and 0.82, respectively). Notably, coherences of "Schizophrenia" and "Major depressive disorder" networks (0.52 and 0.33, respectively) were at the lower end of coherence estimates among all phenotypes. Other psychiatric disorders, such as "Bipolar disorder", "ADHD", showed intermediate coherence estimates (0.76 and 0.55, respectively). In summary, these low coherence estimates highlight the need for further research on the genetics of most cardiovascular and psychiatric diseases to better understand their molecular mechanisms. These results further suggest "Schizophrenia" and "Major depressive disorder" as the top candidates for further genetic studies in psychiatric genome-wide association studies.

@@

>>>>

Network properties, such as connectivity (aka degree distribution) [@Ghiassian:2015aa; @Lee:2008aa; @Barabasi:2004aa; @Feldman:2008aa], can be compared to better understand the relationships among phenotypes [@Vidal:2011aa; @Barabasi:2011aa; @Wang:2011aa]. Networks formed by phenotype-associated genes, or, more precisely, their protein-protein products, are thought to have high connectivity [@Huang:2006aa; @Krauthammer:2004aa; @Goh:2007aa]. This intuition reflects a well-known "guilt-by-association" principle that genes (or, more precisely, their products) that form an interaction network are more likely to share similar functions [@Farber:2013aa], or co-expression patterns [@Huang:2006aa; @Michalak:2008aa]. Although this view has been criticized [@Gillis:2012aa], several studies consistently observed phenotype-associated genes to be either connected as a single network or to participate in common phenotype-specific subnetworks [@Menche:2015aa; @Lee:2011aa; @Akula:2011aa; @Ideker:2008aa; @Iossifov:2008aa]. Therefore, high connectivity among phenotype-specific genes may indicate coherent molecular mechanisms that could be targeted therapeutically. In terms of graph theory, we are asking whether a phenotype-specific network is a community - a highly connected subgraph relatively well-separated from the rest of the network [@fortunato2010community]. However, the direct comparison of connectivity across phenotype-specific networks is hindered by the fact that different phenotypes are associated with different numbers of genes that form networks of different sizes and configurations.

<!--This study compares the connectivity of networks formed by disease- or trait-associated genes and gene products. We hypothesized that genes associated with (categories of) diseases share common molecular mechanisms and, hence, form connected (i.e., coherent) networks. Therefore, the level of connectivity may inform us about similarities and differences among the diseases. The terms “network connectivity,” “coherence,” or “average degree distribution” are used interchangeably as appropriate. We developed a method for normalizing the internal network connectivity that accounts for differences in network sizes formed by disease-associated genes. We compared the slope of normalized internal vs. external network connectivity between psychiatric, neurologic, autoimmune, cancer, cardiovascular, metabolic and other diseases and traits. We used KEGG pathways [30] as a reference for a coherent network with common molecular mechanisms, and we used sets of randomly sampled genes as a reference for the absence of coherence. We found that genes associated with metabolic diseases and traits and autoimmune diseases formed coherent networks similar to KEGG pathways. On the contrary, the coherence of networks of psychiatric and neurologic disease-associated genes, as well as intelligence-associated traits, was similar to that of randomly selected genes. Cancer, anthropometric traits, eye-, and cardiovascular diseases and traits formed networks of intermediate connectivity. Our study identifies significant differences between categories of diseases in terms of network properties of the disease-associated genes and highlights the limitations of the current understanding of the molecular determinants of psychiatric disorders.-->
This study presents a novel measure of network coherence as a slope of the internal versus external degree distributions. The internal degree distribution is defined as gene-specific interaction counts within a protein-protein interaction (PPI) network formed by phenotype-associated genes, while the external degree distribution counts interactions with other genes in the whole PPI network. We used selected gene sets from the MSigDb database [@Subramanian:2005aa] as a reference for highly coherent networks, while sets of randomly sampled genes were used as the reference for the absence of coherence. Using MSigDb and random networks matched in size to phenotype-specific networks, we derived a normalized measure of coherence that can be compared across phenotypes. We hypothesized that the level of coherence may inform us about similarities and differences among phenotypes. Using two PPI databases (STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa]), we compared coherence of 133 phenotypes from GWAScatalog [@Hindorff:2009aa; @MacArthur:2017aa]. Our results show the tendency of complex phenotypes, such as intelligence-related traits, to have low coherence and metabolic traits and diseases to have high coherence. Our method enables direct comparison of coherence measures and highlights gaps in the current understanding of molecular mechanisms of many phenotypes, e.g., Major Depressive Disorder having the lowest coherence in the already low-coherence "Psychiatric disease" category.

@@

To compare coherences, we estimated coherence as the slope of a line through a scatterplot of internal vs. external degree distributions. That is, we visualized internal (X-axis) vs. external (Y-axis) degrees for each phenotype-specific gene on a single plot and fit a regression line through it, enforcing the fit through the origin. A network with the highest coherence (full internal connectivity, zero external connections) would be represented by a horizontal line. Conversely, a network with the lowest coherence (zero internal connectivity, full external connectivity) would be represented by a vertical line. Consequently, slopes of internal vs. external degree distributions of phenotype-specific networks would represent the corresponding levels of coherence. In summary, comparing internal vs. external degree distributions represents a viable metric to quantify and compare the molecular coherence of phenotype-associated networks.

@@

Direct comparison of degree distributions of the networks formed by phenotype-specific genes is hindered by the fact they depend on network size and configuration [@Leskovec:2010:ECA:1772690.1772755]. An intuitive example is to consider a ring and a fully connected network of size 10 and 20 nodes, respectively (Figure 1). The degree distributions of ring networks can be directly compared regardless of network size (vectors of "2" of length 10 and 20, respectively, Figure 1A).  Although the fully connected networks are similar in that they are fully connected (i.e., similarly coherent), their degree distributions differ (all nodes in the 10-node network have a degree "9", while all nodes in the 20-node network have degree "19", Figure 1B). <!--Normalizing the degree distributions of the fully connected networks by dividing them by network size makes them identical as would be expected for networks with similar coherence. However, similar normalization of ring networks will make their degree distributions different (0.22 for the 10-node network and 0.11 for the 20-node network), leading to the illusory impression of differences in coherence between intuitively equally coherent networks. -->These observations highlight the difficulty in comparing degree distributions of phenotype-associated gene networks due to two facts: 1) the network sizes differ, and 2) the network configuration (e.g., fully connected, ring, or intermediate connectivity) is unknown.

@@

Given the dependence of degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we created size-matched references of high and low coherence. Specifically, we selected 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections) and 100 random networks matched in size to the corresponding network of phenotype-specific genes. Consequently, we use median slopes of internal vs. external degree distributions created by these size-matched references to normalize the slopes of phenotype-specific networks to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed alleviates the dependency of coherence on network size (Average PCC = 0.01, Supplementary Figure S1) and allows for direct comparison of normalized coherences across phenotypes.

@@

In addition to size dependency, coherence estimates depend on the choice of the PPI database used to build networks and estimate the corresponding internal vs. external degree distributions. To evaluate the consistency of coherence estimates, we assessed the pairwise correlation of normalized coherences obtained using different PPI databases. Correlation between coherence estimates obtained using "STRING" and "STRING filtered" was high (PCC = 0.80, Supplementary Figure S1). Expectedly, coherence estimates using Biogrid database were less similar to that of "STRING" and "STRING filtered"; however, the overall correlation remained high. Notably, coherence estimates using "STRING filtered" database were most similar to that of "Biogrid" (PCC = 0.58), indicating that filtering step indeed removes "noisy" PPIs and improves quality of coherence estimates. Overall, these results demonstrate consistency in coherence estimates using different PPI databases.

@@

We estimated coherence of 49 traits from five categories (Figure 3, Supplementary Table S3, Supplementary Figures S2-S3). Among anthropometric traits, Weight and Body Mass Index had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, the metabolic trait category had the highest overall coherence (mean coherence 0.86, Table 1), and these results were consistent when using other PPI databases (Supplementary Table S1). Several notable examples of low coherence include Obesity-related traits and Height (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the Intelligence category had low overall coherence (mean coherence = 0.41).

@@

We further estimated coherence of 67 diseases from seven categories (Figure 4, Supplementary Figures S4-S5, Supplementary Table S3). Metabolic diseases had high overall coherence (average coherence 0.91, Table 1), with "Obesity" and "Type 2 diabetes" being the most coherent (1.01 and 0.91, respectively). Similarly, the average coherence for cancer diseases was 0.72, with "Testicular germ cell tumor" and "Pancreatic cancer" being highly coherent (0.92 and 0.74, respectively).

@@

"Rheumatoid arthritis", "Systemic Lupus Erythematosus", and "Systemic sclerosis" were the most coherent in the "Autoimmune" category (0.86, 0.80, 0.79, respectively). On the other end of the coherence spectrum, "Allergic rinitis" and "Atopic dermatitis" were the least coherent (0.51, 0.60, respectively). The average coherence of autoimmune diseases was high, 0.70 (Table 1), with all autoimmune diseases, except "Allergic rinitis", having coherence estimates > 0.60 (Supplementary Table S3). Diseases in the "Neurologic" category were similarly highly coherent (average coherence 0.72). However, the average coherence of neurologic disorders using STRING database was 0.49, suggesting that, with larger number of networks being estimated the overall coherence of neurologic diseases is low. "Alzheimer's disease" and "Parkinson's disease" were the most coherent (0.88 and 0.84, respectively), while "Migraine" was the least coherent (0.54). Diseases in the "Eye disease" category were more heterogeneous, with coherence ranging from 0.35 for "Corneal astigmatism" to 1.01 for "Refractive error". The average coherence of diseases in the "Eye disease" category was relatively high (0.66). 
+-------+
| ![](Figures/Fig4_summary_diseases.png)      |
| **Figure 4. Coherence estimates of diseases.** Size of dots represents the level of normalized coherence (X-axis) for individual diseases (Y-axis).   |
+-------+

@@

Diseases in "Psychiatric" and "Cardiovascular" categories showed the lowest average level of coherence (0.64 and 0.66, respectively, Table 1). The corresponding coherence estimates were highly heterogeneous. Among cardiovascular diseases, "Coronary artery disease" showed the highest level of coherence (0.83), while "Hypertension" had the lowest coherence (0.53). Among psychiatric disorders, "Smoking behavior" and "Alcohol dependence" had high coherence (0.87 and 0.82, respectively). Notably, coherences of "Schizophrenia" and "Major depressive disorder" networks (0.52 and 0.33, respectively) were at the lower end of coherence estimates among all phenotypes. Other psychiatric disorders, such as "Bipolar disorder", "ADHD", showed intermediate coherence estimates (0.76 and 0.55, respectively).

@@

The coherence measure allows us to gain insights into the genetic component of various phenotypes, as revealed by the current state of the corresponding GWASs. The high coherence of metabolic diseases and traits is suggestive of the strong genetic component driving molecular mechanisms of metabolism-specific gene networks. On the contrary, the low coherence of a phenotype suggests either insufficient knowledge of genetics or the true lack of genetic influence. The insufficient knowledge may be remedied by identifying novel genes through larger-scale GWASs and meta-analyses. However, the lack of genetic influence will continue to be reflected in low coherence measures irrespectively of the number of genes. Our results highlight the low coherence of the molecular mechanisms of intelligence-related traits, cardiovascular and psychiatric disorders, but cannot answer whether insufficient knowledge or the lack of genetic component drives it.
